---
title: "RNA-seq Analysis for GSE236155"
author: "Sanny Bold"
date: "24/03/2025"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
All code adapted from the Functional Genomic Technologies (University of Edinburgh, 2026) class materials

## Analysis of RNA-seq data using R 


```{r init, echo=F}
#required libraries & files
library(Rsubread)
library(edgeR)
library(limma)
library(gplots)
library(DESeq2)
library(affy)
library(QuasR)
library(pheatmap)
library(EnhancedVolcano)
library(fgsea)
library(dplyr)
library(ggplot2)
load(file = "mytable_feaures") #file containing alignment and annotations of the RNA-seq experiment for experiment GSE236155 prepared by Prof. Simon Tomlinson


```

```{r input_processing, echo=F}
#access count data and column names.
table_rnaseq<-(mytable_feaures)$counts
colnames_rnaseq <-colnames(table_rnaseq)

#Generate an 'adf data frame' for the annotation.  
adf <-data.frame(Colnames =colnames_rnaseq)
library(stringr)
locc <- str_locate_all(pattern="Aligned.sortedByCoord.out.bam",adf$Colnames)
adf$ShortGEOName <-substr(adf$Colnames,1,locc[[1]][,"start"]-1)

#From the GEO page for this study- https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE236155
#GSM7518952	mouse_NC_1
#GSM7518953	mouse_NC_2
#GSM7518954	mouse_NC_3
#GSM7518955	mouse_OE_1
#GSM7518956	mouse_OE_2
#GSM7518957	mouse_OE_3

#My samples are in the same order as the annotaion so short names created and added to annotation
adf$ShortNames <-c("mouse_NC_1","mouse_NC_2","mouse_NC_3","mouse_OE_1","mouse_OE_2","mouse_OE_3")

#create a factor for the treatment
split_names <-str_split(adf$ShortNames, pattern="_")
treatment <-c()
for(i in 1:length(split_names)){
  treatment <-c(treatment,split_names[[i]][2])
}
adf$Treatment <- treatment

#https://stackoverflow.com/questions/42288075/how-to-autogenerate-color-vector-from-a-numeric-vector
z <- factor(adf$Treatment)
adf$Colours <-rainbow(length(levels(z)))[z]


#ensure colnames of study match the rownames of the annotation
rownames(adf) <-adf$Colnames
if(identical(colnames_rnaseq,rownames(adf))==FALSE){
  print("Names do not match")
}else{
  print("Names match.. OK to proceed")
}

#switch to shortnames to help with the figures

rownames(adf) <-adf$ShortNames
colnames(table_rnaseq) <-adf$ShortNames

colnames_rnaseq <-colnames(table_rnaseq)

```


## Loading the Data into DESeq
DESeq2 will used for the statistical analysis and the normalisation of the data. 
```{r loading_deseq, echo=F}
#Create object for DESeq
dds_rnaseq <- DESeqDataSetFromMatrix(countData = table_rnaseq,colData = adf,design = ~ Treatment)
#check the object dimensions
dim(dds_rnaseq)
head(rownames(dds_rnaseq))
colnames(dds_rnaseq)
#check the object type
class(dds_rnaseq)
typeof(dds_rnaseq)

#check contents
slotNames(dds_rnaseq)
colnames(colData(dds_rnaseq))
colnames(assay(dds_rnaseq))

```

Quality Filtering Genes
Apply filters to remove genes that do not pass quality filters.


```{r apply_filters, echo=F}
#filter based upon counts
keep <- rowSums(counts(dds_rnaseq)) > 1
dds_rnaseq <- dds_rnaseq[keep,]
nrow(dds_rnaseq)
```

## Normalisation of Data

```{r normalisation, echo=F}
#size factor estimation for normalisation-  used to scale by library size when data is exported.
dds_rnaseq <- estimateSizeFactors(dds_rnaseq)
sizeFactors(dds_rnaseq)

```

Multiple normalizations compared. remaining step in the pipeline uses rlog normalization.


```{r plot_normalised, echo=T}

#Normalised log counts
counts_rnaseq <-log2(counts(dds_rnaseq, normalized=TRUE) +1)
#fpm is similar to cpm and discussed here https://rdrr.io/bioc/DESeq2/man/fpm.html.  Here the data is log transformed
fpm_rnaseq <-log(fpm(dds_rnaseq) +1)#this is also normalised by the sizeFactors
#rlog and vsd are discussed here https://genomebiology.biomedcentral.com/articles/10.1186/s13059-014-0550-8
vsd_rnaseq <- vst(dds_rnaseq, blind = T)
head(vsd_rnaseq, 3)
rld_rnaseq <- rlog(dds_rnaseq, blind = T)
head(rld_rnaseq, 3)


```

## Plots with Normalised Data.

```{r build_plots, echo=F}

counts_rnaseq_ <- counts_rnaseq
fpm_rnaseq_ <- fpm_rnaseq
vsd_rnaseq_ <- vsd_rnaseq
rld_rnaseq_ <- rld_rnaseq
colnames(counts_rnaseq_) <- c("NC1","NC2","NC3","OE1","OE2","OE3")
colnames(fpm_rnaseq_) <- c("NC1","NC2","NC3","OE1","OE2","OE3")
colnames(vsd_rnaseq_) <- c("NC1","NC2","NC3","OE1","OE2","OE3")
colnames(rld_rnaseq_) <- c("NC1","NC2","NC3","OE1","OE2","OE3")
#Boxplots
boxplot(counts_rnaseq_, main="log2 RNA-seq Counts Data", xlab = "Samples", ylab="log2 Counts")
boxplot(fpm_rnaseq_, main="RNA-seq Fragments per Million (FPM)", xlab = "Samples", ylab="FPM")
boxplot(assay(vsd_rnaseq_), main="RNA-seq VSD (variance stabilized data)", xlab = "Samples", ylab="VST expression")#etc
boxplot(assay(rld_rnaseq_), main="RNA-seq RLD (rlog transformed data)", xlab = "Samples", ylab="rlog counts")#etc
#but suppose we look at non-normalised
na.rm=T
counts_un <-counts(dds_rnaseq, normalized=FALSE)+1
counts_un_ <- counts_un
colnames(counts_un_) <- c("NC1","NC2","NC3","OE1","OE2","OE3")
boxplot(log2(counts_un_), main="RNA-seq unnormalized Counts Data", xlab = "Samples", ylab="Counts")
counts_un_rlog<-rlog(counts_un)





#mva.pairs plot for data
mva.pairs(counts_un_rlog,na.rm=TRUE, cex=1, main = "MVA plot of samples")

```

Exploratory Data Plots (PCA)

```{r more_explore, echo=F}
#we can run a PCA plot here...
library(scatterplot3d)
library(ggplot2)
# Perform PCA
pca <- prcomp(t(na.omit(assay(rld_rnaseq))), scale=T)
# Plot the PCA results
s3d<-scatterplot3d(pca$x[,1:3], pch=19, color=adf$Colours, main="3D PCA plot for samples")
s3d.coords <- s3d$xyz.convert(pca$x[,1:3])
text(s3d.coords$x, s3d.coords$y, labels = colnames(rld_rnaseq),pos = 3,offset = 0.5,cex=0.5)
#or a quick plot in 2d
qplot(pca$x[,1],pca$x[,2],xlab="PCA1", ylab="PCA2",color=adf$Colours, main = "2D PCA plot for samples")+scale_color_discrete(name="sample", labels=c("mouse_NC", "mouse_OE"))
#We can also output a distance matrix (euclidean distances between rows/smaples)
sampleDists <- dist(t(assay(rld_rnaseq)))
sampleDists

```
## Generate further heatmaps with the data

```{r more_heatmaps, echo=F}
library("pheatmap")
library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
pheatmap(sampleDistMatrix, main = "Heatmap of Sample-to Sample distances")
#easier PCA plot
ress <-plotPCA(rld_rnaseq, intgroup = c("Treatment"))
ress <- ress + geom_label(aes(label = name)) + labs(title = "2D PCA plot of Samples")
plot(ress)
```


## Differential Gene Expression
Simplified stats

```{r differential_expression, echo=F}
dds_rnaseq <- DESeq(dds_rnaseq)

#lfc shrink not applied due to bug in DESeq2 https://support.bioconductor.org/p/77461/#107693
#dds_rnaseq <- lfcShrink(dds_rnaseq, coef=2)




#specify contrast
#note: results(dds, contrast=c("condition","A","B")): genes with logFC > 0 overexpressed in A https://support.bioconductor.org/p/134510/
result_treatment <-results(dds_rnaseq,contrast=c("Treatment","OE","NC"))






#plot of differential expression
plotMA(result_treatment, main="DESeq2 MA plot", ylim=c(-2,2))
table(result_treatment$padj < 0.01)
table(result_treatment$padj < 0.05)
result_treatment_selected <- subset(result_treatment, padj < 0.05)
result_treatment <-result_treatment[order(abs(result_treatment$stat),decreasing = TRUE), ]
head(result_treatment)
```

## Heatmaps from each of the output tables- selected top 50 or all using fold change and FDR.

```{r differential_expression_heatmaps, echo=F}
top70 <-rownames(result_treatment)[1:70]
#make a heatmap for the top 70
pheatmap(assay(rld_rnaseq)[top70,],scale="row",show_rownames=F,main="RLD values, Row Scaled, Top 70")
#all
pheatmap(assay(rld_rnaseq)[rownames(result_treatment),],scale="row",show_rownames=F,main="Treatment Selected, Row Scaled")


```


## Build Gene Annotation. A serialised is generated in case Ensembl is unavailable

```{r build_gene_annotation, echo=F}
#Download Ensembl annotation using BiomaRt and rename the samples
library(biomaRt)
if(file.exists("resultAnnot.RData")){

  load("resultAnnot.RData")
  print("Loaded existing Ensembl annotation...")
}else{
  
  #UK ensembl is being updated, USA mirror used, "useast.ensembl.org"
  ensembl_host <-"https://www.ensembl.org"
  head(biomaRt::listMarts(host = ensembl_host), 15)
  head(biomaRt::listAttributes(biomaRt::useDataset(dataset = "mmusculus_gene_ensembl",mart = useMart("ENSEMBL_MART_ENSEMBL",host = ensembl_host))), 40)
  mart <- biomaRt::useDataset(dataset = "mmusculus_gene_ensembl",mart    = useMart("ENSEMBL_MART_ENSEMBL",host    = ensembl_host))
  resultAnnot <- biomaRt::getBM(values=rownames(dds_rnaseq),attributes =     c("ensembl_gene_id","external_gene_name","chromosome_name","start_position","end_position","description","strand"),filters="ensembl_gene_id",mart=mart)
  save(resultAnnot,file="resultAnnot.RData")
}

```

Merge Annotation with Input Data
```{r merge_gene_annotation, echo=F}
#merge with input data
names <- resultAnnot[,1]
resultAnnot <- as.data.frame(resultAnnot)
resultAnnot <- resultAnnot[!duplicated(resultAnnot$ensembl_gene_id), ]
idx<-match(rownames(dds_rnaseq),resultAnnot$ensembl_gene_id)
resultAnnot <- resultAnnot[idx, ]

unmatched_ids <- rownames(dds_rnaseq)[is.na(resultAnnot$ensembl_gene_id)]
resultAnnot$ensembl_gene_id[is.na(resultAnnot$ensembl_gene_id)] <- unmatched_ids

all(resultAnnot$ensembl_gene_id == rownames(dds_rnaseq))
rownames(resultAnnot) <- resultAnnot$ensembl_gene_id
all(rownames(dds_rnaseq) == rownames(resultAnnot))

grr<-resultAnnot[match(rownames(dds_rnaseq), resultAnnot$ensembl_gene_id),]
all(rownames(dds_rnaseq) == rownames(grr))
resultAnnot <-grr
all(rownames(dds_rnaseq) == rownames(resultAnnot))

#make the nice names
names <- names[idx]
resultAnnot$ensembl_gene_id <- names
nice_names<-ifelse(resultAnnot$external_gene_name == resultAnnot$ensembl_gene_id, 
                   resultAnnot$ensembl_gene_id, 
                   paste(resultAnnot$ensembl_gene_id,resultAnnot$external_gene_name, sep = '_'))
resultAnnot$nice_names <-nice_names
head(resultAnnot)
all(rownames(dds_rnaseq) == rownames(resultAnnot))

#check names
rld_rnaseq <- rlog(dds_rnaseq, blind = TRUE)#
idx2 <-match(rownames(result_treatment)[1:50],rownames(dds_rnaseq))
plotme <-(rld_rnaseq)[rownames(result_treatment)[1:50],]

rownames(plotme)<-resultAnnot$nice_names[idx2]
#make heatmap with candidate genes
png(filename="heatmap_candidates.png",width=800, height=1000)
pheatmap(assay(plotme),scale="row",fontsize_row = 10,cellheight =12, cellwidth =12,treeheight_row = 40, treeheight_col = 40, main = "Row scaled, Top 50")
dev.off()

pheatmap(assay(plotme),scale="row",fontsize_row = 10,cellheight =12, cellwidth =12,treeheight_row = 80, treeheight_col = 100)

```
## Final Results Table


```{r merge_gene_annotation2, echo=F}

#this has the results ordered

#before 
head(result_treatment,50)
result_treatment2 <- result_treatment

annot_2 <- resultAnnot
idx3 <- match(rownames(result_treatment), annot_2$ensembl_gene_id)

new_names <- annot_2$nice_names[idx3]
new_names[is.na(new_names)] <- rownames(result_treatment)[is.na(new_names)]

rownames(result_treatment)<-new_names
#after
print(result_treatment[1:50,])


```

## A volcano plot
a similar plot was made by the authors of the study that generated the RNA-sequence data

```{r make_volcano, echo=F}

result_treatment$gene<-annot_2$external_gene_name[idx3]
result_treatment2$gene <- annot_2$external_gene_name[idx3]
write.csv(as.data.frame(result_treatment), "result_treatment.csv")
top10 <- as.data.frame(result_treatment)
top10 <- top10[order(top10$padj), ]
top10 <- top10[1:10, ]
top10_idx <- match(rownames(top10), resultAnnot$external_gene_name)
top10$ensembl_gene_id <- resultAnnot$ensembl_gene_id[top10_idx]
write.csv(top10, "result_treatment_top10DEGs.csv")
has_annot <- is.na(result_treatment$gene)
has_annot <-!has_annot
result_treatment<-result_treatment[has_annot,]
rownames(result_treatment)<-result_treatment$gene

#Now we have gene names so we can plot the remaining genes in a volcano plot

#these genes were highlighted as interesting in the paper
paper_genes <-c("Cd36","Abca1","Abcg1","Abcg2","Abcg3","Abca8a","Abca8b","Abca9","Pparg", "Msr1")


#just check they're measured in this annotation
if(sum(paper_genes %in% rownames(result_treatment))==length(paper_genes)){
  print("all genes found in data")
}

#for volcano see 
#https://bioconductor.org/packages/devel/bioc/vignettes/EnhancedVolcano/inst/doc/EnhancedVolcano.html
volcano_test <-EnhancedVolcano(result_treatment,
    lab = rownames(result_treatment),
    x = 'log2FoldChange',
    y = 'padj',
    selectLab = paper_genes,
    xlab = bquote(~Log[2]~ 'fold change'),
    ylab = bquote(~-Log[10] ~ "adjusted p-value"),
    subtitle = bquote(italic("p <0.05, |log2FC|≥1")),
    pCutoff = 0.05,
    FCcutoff = 1.0,
    max.overlaps = Inf,
    pointSize = 1.0,
    labSize = 3,
    labCol = 'black',
    labFace = 'bold',
    boxedLabels = TRUE,
    colAlpha = 4/5,
    legendPosition = 'right',
    legendLabSize = 10,
    legendIconSize = 4.0,
    drawConnectors = TRUE,
    widthConnectors = 1.0,
    colConnectors = 'black')

plot(volcano_test,xlim=800,ylim=600)
png("volcano_test.png",width = 800, height = 600)
plot(volcano_test)
dev.off()

print(as.data.frame(result_treatment[paper_genes,]))
write.csv(as.data.frame(result_treatment[paper_genes,]), "paper_genes.csv")


```

```{r paper heatmap}
#note rld_rnaseq <- rlog(dds_rnaseq, blind = T)
papergene_plot <- result_treatment2[result_treatment2$gene %in% paper_genes, ]
papergene_plot <- papergene_plot[match(paper_genes, papergene_plot$gene), ]

papergene_plot$id <- rownames(papergene_plot)
names_paper <- paste(papergene_plot$gene, papergene_plot$id, sep = "_")

rownames(papergene_plot) <- names_paper
papergenes_plotme <- assay(rld_rnaseq[papergene_plot$id, ])
rownames(papergenes_plotme) <- names_paper
pheatmap(papergenes_plotme, scale = "row")
papergene_plot

write.csv(papergene_plot, "paper_genes.csv")
```

## Functional enrichment analysis using gsea

```{r gsea}
results <-read.csv("result_treatment.csv",header =T, row.names = 1)

#fix the rownames
head(results)

results$full_name <- rownames(results)

results$ensembl_gene_id <- sub("_.*$", "", results$full_name)

#grab the hallmark gene set from bioinf.wehi.edu.au
#generally it is better to cache download results, not download each time, but this file is small!
download.file("https://bioinf.wehi.edu.au/software/MSigDB/mouse_H_v5p2.rdata", destfile = "mouse_H_v5p2.rdata")
load("mouse_H_v5p2.rdata")

#We need to map to EntrezIDs- the annotation uses EntrezID
head(Mm.H)

#
library(biomaRt)
mart <- useDataset("mmusculus_gene_ensembl", mart=useMart("ensembl"))
#listAttributes(mart)
ens2entrez <- getBM(attributes=c("ensembl_gene_id", "entrezgene_id"), mart=mart)
head(ens2entrez)

#Map gene names to entrezgene_id
#if NA remove (we only need to consider mappable genes)

results2 <- inner_join(results, ens2entrez, by=join_by("ensembl_gene_id"))
results3 <- results2

head(results2,n=40)

#sanity check one or two of the results eg
#ENSMUSG00000046101  303.147520    -0.26407455 0.1563231 -1.68928724 0.0911643983 0.277178884        240697
ens2entrez[which(ens2entrez$ensembl_gene_id  == "ENSMUSG00000046101"),]

#remove any that are NA
results2 <-results2[is.na(results2$entrezgene_id)==FALSE,]
dim(results2)

head(results2)

#This analysis takes a ranked list of numbers with entrez IDs as names
#We need to choose how to order the hits
#Here I order by stat value
results2 <-results2[order(results2$stat,decreasing=T),]

results2 <-results2[,c("stat","entrezgene_id")]

#remove non-unique row names
results2 <-results2[!duplicated(results2$entrezgene_id),]


rownames(results2)<-results2[,"entrezgene_id"]
results2["entrezgene_id"] <-NULL


#it might want a named vector
rn <-rownames(results2)
colnames(results2) <-NULL
results2 <-results2[,1]
names(results2)<-rn

head(results2)

#perform the enrichment analysis
results2  <- results2[!(is.na(names(results2)))]
results2  <- results2[!(is.na(results2))]
#results2 <-log2(results2)

fgseaRes <- fgsea(Mm.H, results2, minSize=25, maxSize = 500)

fgseaRes <-fgseaRes[order(fgseaRes$NES,decreasing=T),]
head(fgseaRes, 50)

plotEnrichment(Mm.H[["HALLMARK_FATTY_ACID_METABOLISM"]],results2) +labs(title="HALLMARK FATTY ACID METABOLISM")
plotEnrichment(Mm.H[["HALLMARK_MYOGENESIS"]],results2) +labs(title="HALLMARK MYOGENESIS")
plotEnrichment(Mm.H[["HALLMARK_INFLAMMATORY_RESPONSE"]],results2) +labs(title="HALLMARK INFLAMMATORY RESPONSE")


#plotGseatable, adapted from: https://bioconductor.org/packages/release/bioc/vignettes/fgsea/inst/doc/fgsea-tutorial.html
toppathwaysup <- fgseaRes[NES>0][head(order(padj), n=10), pathway]
toppathwaysdown <- fgseaRes[NES<0][head(order(padj), n=10), pathway]
toppathways <- c(toppathwaysup, rev(toppathwaysdown))

plotGseaTable(Mm.H[toppathways], results2, fgseaRes, gseaParam=0.5)+labs(title="Top pathways")

collapsedPathways <- collapsePathways(fgseaRes[order(padj)][padj < 0.05], 
                                      Mm.H, results2)
mainPathways <- fgseaRes[pathway %in% collapsedPathways$mainPathways][
  order(-NES), pathway]

png("cppw.png", height = 5000, width = 5000, res = 250)
plotGseaTable(Mm.H[mainPathways], results2, fgseaRes, 
              gseaParam = 0.5)+labs(title="Top collapsed pathways")
dev.off()

#barplot of top pathways
idx_pw <- match(toppathways, fgseaRes$pathway)
df <- data_frame(x=toppathways, y=fgseaRes$NES[idx_pw])
df <- df[order(df$y), ]

ggplot(df, aes(x=x, y=y))+
  geom_col()+
  labs(
    title = "Top pathways (up and down)",
    x="pathways",
    y="NES"
  )+
  coord_flip()

#investigating Myogenesis genes
myo_leads <- fgseaRes$leadingEdge[fgseaRes$pathway == "HALLMARK_MYOGENESIS"]
myo_leads <- unlist(myo_leads)
myos <- results3[as.character(results3$entrezgene_id) %in% myo_leads, ]

write.csv(myos, "myogenes.csv")

```

## Session Information

```{r session_info, echo=F}

sessionInfo()

```

